---
title: "STRUCTURE Individual-Level Mean Barplots Across K"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(clue)

report_dir <- dirname(knitr::current_input(dir = TRUE))
repo_root <- normalizePath(file.path(report_dir, "..", ".."), mustWork = FALSE)

structure_dir <- file.path(repo_root, "results", "structureHarvester")
sample_file <- file.path(repo_root, "data", "PG_All_structure.txt")
output_dir <- file.path(repo_root, "figure", "STRUCTURE_individual_allK")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## Goal

This report uses the `structureHarvester` `K*.indfile` outputs and the sample order from `PG_All_structure.txt` to build individual-level STRUCTURE bar plots for `K = 1` to `K = 13`.

Workflow highlights:

- each `K*.indfile` contains 5 replicate runs
- cluster labels are aligned across runs using a one-to-one assignment step
- the 5 aligned runs are averaged for each individual
- individuals are shown in the original sample order, nested by population within state groups

## Read Sample Metadata and Order

```{r sample-metadata}
sample_lines <- readr::read_lines(sample_file)[-1]

sample_meta <- tibble(raw_line = sample_lines) %>%
  mutate(
    parts = str_split(str_squish(raw_line), "\\s+"),
    sample_id = map_chr(parts, 1),
    population = str_extract(sample_id, "Pg\\d+"),
    state = case_when(
      population == "Pg346" ~ "IN",
      population %in% c("Pg347", "Pg349", "Pg350", "Pg351", "Pg352", "Pg354") ~ "KY",
      population %in% c("Pg348", "Pg353", "Pg355") ~ "EX-S",
      population %in% c("Pg356", "Pg357") ~ "TNE",
      population %in% c("Pg358", "Pg359", "Pg360", "Pg361", "Pg362") ~ "TNW",
      TRUE ~ "Unknown"
    ),
    sample_order = row_number()
  ) %>%
  select(sample_id, population, state, sample_order)

sample_meta
```

## Helper Functions

```{r helpers}
read_indfile <- function(path, k) {
  lines <- readr::read_lines(path) %>%
    keep(~ str_detect(.x, ":"))

  parsed <- purrr::map_dfr(lines, function(line) {
    line <- str_squish(line)
    pieces <- str_split_fixed(line, " : ", 2)
    left <- str_split(pieces[1], "\\s+")[[1]]
    qvals <- as.numeric(str_split(pieces[2], "\\s+")[[1]])

    tibble(
      individual_index = as.integer(left[1]),
      pop_code = as.integer(left[length(left)]),
      q = list(qvals)
    )
  })

  qmat <- do.call(rbind, parsed$q)
  colnames(qmat) <- paste0("cluster_", seq_len(k))

  bind_cols(parsed %>% select(individual_index, pop_code), as_tibble(qmat))
}

align_and_average_runs <- function(ind_tbl, sample_meta, k, n_runs = 5) {
  n_samples <- nrow(sample_meta)
  stopifnot(nrow(ind_tbl) == n_samples * n_runs)

  run_list <- purrr::map(seq_len(n_runs), function(run_id) {
    start_idx <- (run_id - 1) * n_samples + 1
    end_idx <- run_id * n_samples
    ind_tbl[start_idx:end_idx, paste0("cluster_", seq_len(k))] %>%
      as.matrix()
  })

  ref <- run_list[[1]]
  aligned <- list(ref)

  if (k > 1) {
    for (run_id in 2:n_runs) {
      current <- run_list[[run_id]]
      similarity <- cor(ref, current, use = "pairwise.complete.obs")
      similarity[is.na(similarity)] <- -1
      similarity <- similarity - min(similarity) + 1e-9
      perm <- solve_LSAP(similarity, maximum = TRUE)
      aligned[[run_id]] <- current[, perm, drop = FALSE]
    }
  } else {
    for (run_id in 2:n_runs) {
      aligned[[run_id]] <- run_list[[run_id]]
    }
  }

  avg_q <- Reduce(`+`, aligned) / n_runs
  colnames(avg_q) <- paste0("cluster_", seq_len(k))

  bind_cols(sample_meta, as_tibble(avg_q))
}

build_barplot <- function(avg_tbl, k) {
  cluster_cols <- paste0("cluster_", seq_len(k))

  long_tbl <- avg_tbl %>%
    pivot_longer(cols = all_of(cluster_cols), names_to = "cluster", values_to = "membership") %>%
    mutate(cluster = factor(cluster, levels = cluster_cols), x = sample_order)

  pop_breaks <- avg_tbl %>%
    group_by(population, state) %>%
    summarise(
      xmin = min(sample_order) - 0.5,
      xmax = max(sample_order) + 0.5,
      xmid = mean(sample_order),
      .groups = "drop"
    )

  state_breaks <- avg_tbl %>%
    group_by(state) %>%
    summarise(
      xmin = min(sample_order) - 0.5,
      xmax = max(sample_order) + 0.5,
      xmid = mean(sample_order),
      .groups = "drop"
    )

  pop_separators <- head(pop_breaks$xmax, -1)
  state_separators <- head(state_breaks$xmax, -1)

  cluster_palette <- grDevices::hcl.colors(k, palette = "Set 3", rev = FALSE)
  names(cluster_palette) <- cluster_cols

  ggplot(long_tbl, aes(x = x, y = membership, fill = cluster)) +
    geom_col(width = 1, color = NA) +
    geom_vline(xintercept = pop_separators, color = "white", linewidth = 0.25) +
    geom_vline(xintercept = state_separators, color = "gray20", linewidth = 0.55) +
    annotate("text", x = pop_breaks$xmid, y = -0.035, label = pop_breaks$population, angle = 90, size = 2.1, vjust = 0.5) +
    annotate("text", x = state_breaks$xmid, y = 1.04, label = state_breaks$state, size = 3.2, fontface = "bold") +
    scale_fill_manual(values = cluster_palette, name = "Cluster") +
    scale_x_continuous(expand = c(0, 0), breaks = NULL) +
    scale_y_continuous(limits = c(-0.06, 1.08), breaks = seq(0, 1, by = 0.2), expand = c(0, 0)) +
    coord_cartesian(clip = "off") +
    labs(
      title = paste0("STRUCTURE ancestry composition at K = ", k),
      subtitle = "Individual-level mean ancestry across 5 aligned runs; populations nested within state groups",
      x = NULL,
      y = "Mean ancestry coefficient"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      legend.position = "right",
      plot.margin = margin(15, 15, 40, 15),
      plot.title = element_text(face = "bold")
    )
}
```

## Example: K = 11

```{r k11-example, fig.width=14, fig.height=5.8}
k <- 11
ind_tbl <- read_indfile(file.path(structure_dir, paste0("K", k, ".indfile")), k)
avg_tbl <- align_and_average_runs(ind_tbl, sample_meta, k = k, n_runs = 5)
build_barplot(avg_tbl, k)
```

## Interpretation Notes

The averaged plots should be read with two points in mind:

- the alignment step reduces label switching across runs by matching clusters to the first run before averaging
- higher K values may still show more instability and finer subdivisions even after alignment, so they should be interpreted together with the Evanno and LnP summaries

In this dataset:

- `K = 2` is the strongest Evanno-supported upper-level split
- larger K values likely reflect progressively finer hierarchical structure within `P. globosa`
- the nested layout helps show whether those finer components are mainly state-level, population-level, or admixed within populations
```
