---
title: "Physaria Table 5 Heatmap Workflow"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)

report_dir <- dirname(knitr::current_input(dir = TRUE))
repo_root <- normalizePath(file.path(report_dir, "..", ".."), mustWork = FALSE)

input_file <- file.path(repo_root, "data", "Table_5.csv")
output_png <- file.path(repo_root, "figure", "Table_5_pairwise_FST_GST_heatmap.png")
```

## Goal

This report reads `Table_5.csv`, separates the upper and lower matrix triangles into two metrics, and plots them together as a single heatmap:

- upper diagonal: pairwise `FST`
- lower diagonal: pairwise `GST`

The workflow keeps the data in a tidy format before plotting so the result is easy to modify later.

## Read the Raw Table

```{r read-table}
raw_lines <- readr::read_lines(input_file)
data_lines <- raw_lines[!str_detect(raw_lines, '^"?#') & raw_lines != ""]

table_5 <- readr::read_csv(
  I(data_lines),
  na = c("", "NA"),
  show_col_types = FALSE
)

population_col <- names(table_5)[1]
table_5 <- table_5 %>%
  rename(population = all_of(population_col))

table_5
```

## Confirm Population Order

The row order is important because values above the diagonal become `FST`, while values below the diagonal become `GST`.

```{r population-order}
population_levels <- table_5$population
population_levels
```

## Convert the Matrix to Tidy Format

```{r tidy-format}
table_5_long <- table_5 %>%
  pivot_longer(
    cols = -population,
    names_to = "population_2",
    values_to = "value"
  ) %>%
  mutate(
    population_1 = population,
    row_id = match(population_1, population_levels),
    col_id = match(population_2, population_levels),
    metric = case_when(
      row_id < col_id ~ "FST",
      row_id > col_id ~ "GST",
      TRUE ~ "Diagonal"
    )
  )

table_5_long
```

## Remove the Diagonal and Missing Cells

```{r filter-cells}
heatmap_tbl <- table_5_long %>%
  filter(metric != "Diagonal", !is.na(value)) %>%
  transmute(
    metric,
    population_1 = factor(population_1, levels = rev(population_levels)),
    population_2 = factor(population_2, levels = population_levels),
    value
  )

heatmap_tbl
```

## Check Metric Counts and Ranges

```{r metric-summary}
heatmap_tbl %>%
  group_by(metric) %>%
  summarise(
    n_cells = n(),
    min_value = min(value),
    max_value = max(value),
    mean_value = mean(value),
    .groups = "drop"
  )
```

## Plot the Heatmap

```{r heatmap, fig.width=11, fig.height=10}
heatmap_plot <- ggplot(heatmap_tbl, aes(x = population_2, y = population_1, fill = value)) +
  geom_tile(color = "white", linewidth = 0.35) +
  geom_text(aes(label = sprintf("%.3f", value)), size = 2.3) +
  scale_fill_viridis_c(
    option = "magma",
    name = "Genetic distance"
  ) +
  labs(
    title = "Pairwise genetic distance heatmap",
    subtitle = "Upper diagonal: FST | Lower diagonal: GST",
    x = NULL,
    y = NULL
  ) +
  coord_fixed() +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.text.y = element_text(size = 8),
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

heatmap_plot
```

## Save the Figure

```{r save-output}
ggsave(output_png, plot = heatmap_plot, width = 11, height = 10, dpi = 600)
output_png
```
