---
title: "Physaria Table 3 Microsatellite Summary Workflow"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggrepel)

report_dir <- dirname(knitr::current_input(dir = TRUE))
repo_root <- normalizePath(file.path(report_dir, "..", ".."), mustWork = FALSE)

input_file <- file.path(repo_root, "data", "Table_3.csv")
table_4_file <- file.path(repo_root, "data", "Table_4.csv")
output_dir <- file.path(repo_root, "figure", "Table_3")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## Goal

This report parses `Table_3.csv` and integrates the bottleneck test results from `Table_4.csv` into a set of useful plots for microsatellite summary statistics. The goal is to show:

- diversity patterns across populations
- observed versus expected heterozygosity
- fixation and null-allele diagnostics
- relative metric highs and lows across populations
- species-level averages reported in the manuscript
- Table 4 bottleneck significance alongside heterozygosity patterns

## Read and Parse Table 3

```{r read-table}
raw_lines <- readr::read_lines(input_file)
data_lines <- raw_lines[!str_detect(raw_lines, '^"?#') & raw_lines != ""]

table_3 <- readr::read_csv(I(data_lines), na = c("", "NA"), show_col_types = FALSE) %>%
  mutate(across(c(n, A, AR, AP, Ho, He, F, `% null`, FB), as.numeric)) %>%
  mutate(
    row_type = case_when(
      str_detect(Population, "^Average across") ~ "average",
      if_all(all_of(c("n", "A", "AR", "AP", "Ho", "He", "F", "% null", "FB")), is.na) ~ "section",
      TRUE ~ "population"
    ),
    species_header = if_else(row_type == "section", Population, NA_character_)
  ) %>%
  fill(species_header, .direction = "down") %>%
  mutate(
    species = case_when(
      str_detect(Population, "^Average across P\\. globosa") ~ "P. globosa",
      str_detect(Population, "^Average across P\\. ouachitensis") ~ "P. ouachitensis",
      str_detect(Population, "^Average across P\\. filiformis") ~ "P. filiformis",
      str_detect(Population, "^Average across P\\. gracilis") ~ "P. gracilis",
      TRUE ~ species_header
    )
  ) %>%
  select(-species_header)

table_3
```

## Split Population Rows and Average Rows

```{r split-rows}
extract_significance <- function(x) {
  case_when(
    str_detect(x, fixed("***")) ~ "P < 0.0001",
    str_detect(x, fixed("**")) ~ "P < 0.001",
    str_detect(x, fixed("*")) ~ "P < 0.01",
    TRUE ~ "Not significant"
  )
}

parse_stat_value <- function(x) {
  readr::parse_number(x)
}

table_4_lines <- readr::read_lines(table_4_file)
table_4_data_lines <- table_4_lines[!str_detect(table_4_lines, '^"?#') & table_4_lines != ""]

table_4 <- readr::read_csv(I(table_4_data_lines), show_col_types = FALSE) %>%
  transmute(
    Population,
    `Geographic area`,
    het_excess_z_raw = `Excess in heterozygosity Z-test`,
    het_excess_w_raw = `Excess in heterozygosity Wilcoxon signed-rank test`,
    mratio_z_raw = `Deficiency in M-Ratio Z-test`,
    mratio_w_raw = `Deficiency in M-Ratio Wilcoxon signed-rank test`,
    het_excess_z = parse_stat_value(het_excess_z_raw),
    het_excess_w = parse_stat_value(het_excess_w_raw),
    mratio_z = parse_stat_value(mratio_z_raw),
    mratio_w = parse_stat_value(mratio_w_raw),
    het_excess_z_sig = extract_significance(het_excess_z_raw),
    het_excess_w_sig = extract_significance(het_excess_w_raw),
    mratio_z_sig = extract_significance(mratio_z_raw),
    mratio_w_sig = extract_significance(mratio_w_raw)
  ) %>%
  mutate(
    het_bottleneck_sig = case_when(
      het_excess_z_sig == "P < 0.0001" | het_excess_w_sig == "P < 0.0001" ~ "P < 0.0001",
      het_excess_z_sig == "P < 0.001" | het_excess_w_sig == "P < 0.001" ~ "P < 0.001",
      het_excess_z_sig == "P < 0.01" | het_excess_w_sig == "P < 0.01" ~ "P < 0.01",
      TRUE ~ "Not significant"
    ),
    mratio_bottleneck_sig = case_when(
      mratio_z_sig == "P < 0.0001" | mratio_w_sig == "P < 0.0001" ~ "P < 0.0001",
      mratio_z_sig == "P < 0.001" | mratio_w_sig == "P < 0.001" ~ "P < 0.001",
      mratio_z_sig == "P < 0.01" | mratio_w_sig == "P < 0.01" ~ "P < 0.01",
      TRUE ~ "Not significant"
    )
  )

population_tbl <- table_3 %>% filter(row_type == "population")
average_tbl <- table_3 %>% filter(row_type == "average")

population_tbl <- population_tbl %>%
  left_join(table_4, by = c("Population", "Geographic area")) %>%
  mutate(
    het_bottleneck_sig = replace_na(het_bottleneck_sig, "Not tested"),
    mratio_bottleneck_sig = replace_na(mratio_bottleneck_sig, "Not tested")
  )

population_tbl
```

## Suggested Plot Types for This Table

These are the most informative plot families for this dataset:

- faceted diversity plots for `A`, `AR`, and `AP`
- heterozygosity profile plots comparing `Ho` and `He`
- `Ho` versus `He` scatterplots with a 1:1 reference line
- bottleneck significance overlays from Table 4 on the heterozygosity plots
- fixation/null diagnostics for `F`, `FB`, and `% null`
- a dedicated bottleneck-statistics panel for P. globosa
- standardized heatmaps across all metrics
- species-level average summaries from the manuscript

## Diversity Metrics by Population

```{r diversity-plot, fig.width=10, fig.height=11}
species_colors <- c(
  "P. globosa" = "#1f78b4",
  "P. ouachitensis" = "#e31a1c",
  "P. filiformis" = "#33a02c",
  "P. gracilis" = "#ff7f00"
)

sig_shapes <- c(
  "Not tested" = 21,
  "Not significant" = 21,
  "P < 0.01" = 24,
  "P < 0.001" = 22,
  "P < 0.0001" = 23
)

sig_fills <- c(
  "Not tested" = "white",
  "Not significant" = "white",
  "P < 0.01" = "#fdd835",
  "P < 0.001" = "#fb8c00",
  "P < 0.0001" = "#d7301f"
)

metric_labels <- c(
  A = "Alleles per locus (A)",
  AR = "Allelic richness (AR)",
  AP = "Private alleles (AP)",
  Ho = "Observed heterozygosity (Ho)",
  He = "Expected heterozygosity (He)",
  F = "Fixation index (F)",
  `% null` = "Null alleles (%)",
  FB = "Adjusted fixation index (FB)"
)

diversity_long <- population_tbl %>%
  select(Population, species, A, AR, AP) %>%
  pivot_longer(cols = c(A, AR, AP), names_to = "metric", values_to = "value") %>%
  mutate(
    metric = factor(metric, levels = c("A", "AR", "AP"), labels = metric_labels[c("A", "AR", "AP")]),
    Population = fct_reorder(Population, value)
  )

ggplot(diversity_long, aes(x = value, y = Population, color = species)) +
  geom_segment(aes(x = 0, xend = value, y = Population, yend = Population), linewidth = 0.35, alpha = 0.5) +
  geom_point(size = 2.2) +
  facet_wrap(~metric, scales = "free_x", ncol = 1) +
  scale_color_manual(values = species_colors) +
  labs(title = "Microsatellite diversity metrics by population", x = NULL, y = NULL, color = "Species") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank(), legend.position = "bottom")
```

## Heterozygosity Patterns

```{r heterozygosity-profile, fig.width=11, fig.height=8.5}
het_long <- population_tbl %>%
  mutate(Population = fct_reorder(Population, He)) %>%
  select(Population, species, Ho, He, het_bottleneck_sig) %>%
  pivot_longer(cols = c(Ho, He), names_to = "metric", values_to = "value") %>%
  mutate(
    metric = recode(metric, Ho = "Observed (Ho)", He = "Expected (He)")
  )

ggplot(het_long, aes(x = value, y = Population, color = metric)) +
  geom_line(aes(group = Population), color = "gray70", linewidth = 0.4) +
  geom_point(aes(shape = het_bottleneck_sig, fill = het_bottleneck_sig), size = 2.6, stroke = 0.35) +
  facet_wrap(~species, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c("Observed (Ho)" = "#1b9e77", "Expected (He)" = "#d95f02")) +
  scale_shape_manual(values = sig_shapes, drop = FALSE) +
  scale_fill_manual(values = sig_fills, drop = FALSE) +
  labs(
    title = "Observed versus expected heterozygosity",
    subtitle = "Point symbols show heterozygosity-excess bottleneck significance from Table 4",
    x = "Heterozygosity",
    y = NULL,
    color = NULL,
    shape = "Het. excess test",
    fill = "Het. excess test"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r heterozygosity-scatter, fig.width=9, fig.height=7}
ggplot(population_tbl, aes(x = He, y = Ho, color = species, size = n)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray55") +
  geom_point(aes(shape = het_bottleneck_sig, fill = het_bottleneck_sig), alpha = 0.95, stroke = 0.4) +
  geom_text_repel(aes(label = Population), size = 2.7, max.overlaps = 50, show.legend = FALSE) +
  scale_color_manual(values = species_colors) +
  scale_size_continuous(range = c(2, 7)) +
  scale_shape_manual(values = sig_shapes, drop = FALSE) +
  scale_fill_manual(values = sig_fills, drop = FALSE) +
  coord_equal() +
  labs(
    title = "Observed versus expected heterozygosity across populations",
    subtitle = "Points below the 1:1 line indicate heterozygote deficiency; symbols show Table 4 heterozygosity-excess significance",
    x = "Expected heterozygosity (He)",
    y = "Observed heterozygosity (Ho)",
    color = "Species",
    size = "Sample size",
    shape = "Het. excess test",
    fill = "Het. excess test"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## Fixation and Null Diagnostics

```{r diagnostics, fig.width=10, fig.height=11}
diagnostic_long <- population_tbl %>%
  select(Population, species, F, FB, `% null`) %>%
  pivot_longer(cols = c(F, FB, `% null`), names_to = "metric", values_to = "value") %>%
  mutate(
    metric = factor(metric, levels = c("F", "FB", "% null"), labels = metric_labels[c("F", "FB", "% null")]),
    Population = fct_reorder(Population, value)
  )

ggplot(diagnostic_long, aes(x = value, y = Population, color = species)) +
  geom_vline(xintercept = 0, color = "gray80", linewidth = 0.4) +
  geom_point(size = 2) +
  facet_wrap(~metric, scales = "free_x", ncol = 1) +
  scale_color_manual(values = species_colors) +
  labs(title = "Fixation and null-allele diagnostics", x = NULL, y = NULL, color = "Species") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank(), legend.position = "bottom")
```

## Standardized Metric Heatmap

## Table 4 Bottleneck Statistics

```{r bottleneck-panel, fig.width=10, fig.height=11}
bottleneck_long <- population_tbl %>%
  filter(species == "P. globosa", !is.na(het_excess_z)) %>%
  select(
    Population,
    `Geographic area`,
    het_excess_z,
    het_excess_w,
    mratio_z,
    mratio_w,
    het_excess_z_sig,
    het_excess_w_sig,
    mratio_z_sig,
    mratio_w_sig
  ) %>%
  pivot_longer(
    cols = c(het_excess_z, het_excess_w, mratio_z, mratio_w),
    names_to = "test",
    values_to = "statistic"
  ) %>%
  mutate(
    sig = case_when(
      test == "het_excess_z" ~ het_excess_z_sig,
      test == "het_excess_w" ~ het_excess_w_sig,
      test == "mratio_z" ~ mratio_z_sig,
      test == "mratio_w" ~ mratio_w_sig
    ),
    test = factor(
      test,
      levels = c("het_excess_z", "het_excess_w", "mratio_z", "mratio_w"),
      labels = c(
        "Het. excess Z-test",
        "Het. excess Wilcoxon",
        "M-ratio deficiency Z-test",
        "M-ratio deficiency Wilcoxon"
      )
    ),
    Population = fct_reorder(Population, statistic)
  )

ggplot(bottleneck_long, aes(x = statistic, y = Population)) +
  geom_vline(xintercept = 0, color = "gray82", linewidth = 0.4) +
  geom_segment(aes(x = 0, xend = statistic, y = Population, yend = Population), color = "gray75", linewidth = 0.35) +
  geom_point(aes(fill = sig, shape = sig), color = "black", size = 2.6, stroke = 0.3) +
  facet_wrap(~test, scales = "free_x", ncol = 1) +
  scale_shape_manual(values = sig_shapes[c("Not significant", "P < 0.01", "P < 0.001", "P < 0.0001")], drop = FALSE) +
  scale_fill_manual(values = sig_fills[c("Not significant", "P < 0.01", "P < 0.001", "P < 0.0001")], drop = FALSE) +
  labs(
    title = "Table 4 bottleneck test statistics for P. globosa",
    subtitle = "Symbols encode reported significance levels",
    x = "Test statistic",
    y = NULL,
    shape = "Significance",
    fill = "Significance"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank(), legend.position = "bottom")
```

## Standardized Metric Heatmap

```{r heatmap, fig.width=11, fig.height=8.5}
heatmap_long <- population_tbl %>%
  select(Population, species, A, AR, AP, Ho, He, F, `% null`, FB) %>%
  pivot_longer(cols = c(A, AR, AP, Ho, He, F, `% null`, FB), names_to = "metric", values_to = "value") %>%
  group_by(metric) %>%
  mutate(z_value = as.numeric(scale(value))) %>%
  ungroup() %>%
  mutate(
    metric = factor(metric, levels = c("A", "AR", "AP", "Ho", "He", "F", "% null", "FB"), labels = metric_labels[c("A", "AR", "AP", "Ho", "He", "F", "% null", "FB")])
  )

ggplot(heatmap_long, aes(x = metric, y = Population, fill = z_value)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0, name = "Z-score") +
  labs(
    title = "Standardized microsatellite summary heatmap",
    subtitle = "Values are scaled within each metric to highlight relative highs and lows",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))
```

## Species-Level Averages

```{r species-averages, fig.width=12, fig.height=8}
average_long <- average_tbl %>%
  select(species, A, AR, AP, Ho, He, F, `% null`, FB) %>%
  pivot_longer(cols = -species, names_to = "metric", values_to = "value") %>%
  mutate(
    metric = factor(metric, levels = c("A", "AR", "AP", "Ho", "He", "F", "% null", "FB"), labels = metric_labels[c("A", "AR", "AP", "Ho", "He", "F", "% null", "FB")])
  )

ggplot(average_long, aes(x = species, y = value, fill = species)) +
  geom_col(width = 0.72) +
  facet_wrap(~metric, scales = "free_y", ncol = 4) +
  scale_fill_manual(values = species_colors, guide = "none") +
  labs(title = "Species-level averages reported in Table 3", x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```
