---
title: "Physaria Table 1 Map Workflow"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(ggrepel)

input_file <- "../data/Table_1.csv"
output_png <- "../figure/Table_1_map.png"
```

## Goal

This report rebuilds a publication-style map from `Table_1.csv` using the same core package stack and plotting style as the reference script:

- `tidyverse`
- `sf`
- `rnaturalearth`
- `rnaturalearthdata`
- `ggspatial`
- `ggrepel`

It also exposes the intermediate objects so the map can be tuned later.

## Read and Clean Table 1

```{r read-clean}
raw_lines <- readr::read_lines(input_file)
data_lines <- raw_lines[!str_detect(raw_lines, '^"?#') & raw_lines != ""]

table_1 <- readr::read_csv(I(data_lines), show_col_types = FALSE) %>%
  rename(
    population_code = `Population code`,
    collector = Collector,
    species_name = `Species name`,
    locality = `State, County, Locality (Element Occurrence number)`,
    ecoregion = Ecoregion,
    lat_raw = Lat,
    long_raw = Long,
    n_samples = `Number of samples`
  ) %>%
  mutate(
    species_name = str_replace_all(species_name, "[“”]", ""),
    species_name = str_squish(species_name),
    lat = readr::parse_number(as.character(lat_raw)),
    long = readr::parse_number(as.character(long_raw)),
    n_samples = readr::parse_number(as.character(n_samples)),
    species_group = case_when(
      str_detect(species_name, "globosa") ~ "P. globosa",
      str_detect(species_name, "ouachitensis") ~ "P. ouachitensis",
      str_detect(species_name, "filiformis") ~ "P. filiformis",
      str_detect(species_name, "gracilis") ~ "P. gracilis",
      TRUE ~ species_name
    )
  )

table_1
```

## Check Coordinate Parsing

This is mainly for debugging malformed coordinates from the extracted CSV. `parse_number()` is used so entries such as `36.501393, -` still resolve to usable numeric latitude values.

```{r coord-check}
table_1 %>%
  transmute(population_code, lat_raw, long_raw, lat, long) %>%
  filter(is.na(lat) | is.na(long) | str_detect(as.character(lat_raw), ",") | str_detect(as.character(long_raw), ",")) 
```

## Build Spatial Objects

```{r spatial-objects}
site_tbl <- table_1 %>%
  filter(!is.na(lat), !is.na(long))

site_sf <- st_as_sf(site_tbl, coords = c("long", "lat"), crs = 4326, remove = FALSE)
site_sf_3857 <- st_transform(site_sf, 3857)

label_sf <- site_tbl %>%
  group_by(population_code, species_group) %>%
  summarise(
    long = mean(long),
    lat = mean(lat),
    n_samples = sum(n_samples),
    .groups = "drop"
  ) %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326) %>%
  st_transform(3857) %>%
  mutate(label = paste0(population_code, " (n=", n_samples, ")"))

label_xy <- cbind(st_drop_geometry(label_sf), st_coordinates(label_sf))

usa_state <- ne_states(country = "United States of America", returnclass = "sf") %>%
  st_transform(3857)

bbox <- st_bbox(site_sf_3857)
x_pad <- (bbox$xmax - bbox$xmin) * 0.18
y_pad <- (bbox$ymax - bbox$ymin) * 0.18

list(
  n_points = nrow(site_tbl),
  bbox_3857 = bbox
)
```

## Base Map

```{r base-map, fig.width=10, fig.height=8}
ggplot() +
  geom_sf(data = usa_state, fill = NA, color = "gray75", linewidth = 0.3) +
  coord_sf(
    xlim = c(bbox$xmin - x_pad, bbox$xmax + x_pad),
    ylim = c(bbox$ymin - y_pad, bbox$ymax + y_pad),
    expand = FALSE,
    crs = 3857
  ) +
  theme_bw() +
  theme(panel.grid = element_blank())
```

## Add Sampling Points

```{r point-map, fig.width=10, fig.height=8}
fill_values <- c(
  "P. globosa" = "#c44e52",
  "P. ouachitensis" = "#4c78a8",
  "P. filiformis" = "#55a868",
  "P. gracilis" = "#8172b3"
)

point_map <- ggplot() +
  geom_sf(data = usa_state, fill = NA, color = "gray75", linewidth = 0.3) +
  geom_sf(
    data = site_sf_3857,
    aes(fill = species_group, size = n_samples),
    shape = 21,
    color = "black",
    stroke = 0.3,
    alpha = 0.9
  ) +
  scale_fill_manual(values = fill_values, drop = FALSE, name = "Species") +
  scale_size_continuous(range = c(2.2, 5.5), name = "Sample size") +
  coord_sf(
    xlim = c(bbox$xmin - x_pad, bbox$xmax + x_pad),
    ylim = c(bbox$ymin - y_pad, bbox$ymax + y_pad),
    expand = FALSE,
    crs = 3857
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.box = "vertical"
  )

point_map
```

## Add Labels and Scale Bar

```{r final-map, fig.width=10, fig.height=8}
final_map <- point_map +
  geom_text_repel(
    data = label_xy,
    aes(x = X, y = Y, label = label),
    size = 3,
    min.segment.length = 0.1,
    seed = 42,
    max.overlaps = 50,
    box.padding = 0.25,
    point.padding = 0.2,
    segment.color = "gray40"
  ) +
  annotation_scale(location = "br", width_hint = 0.22) +
  labs(x = NULL, y = NULL) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

final_map
```

## Save the Final PNG

```{r save-output}
ggsave(output_png, plot = final_map, width = 10, height = 8, dpi = 900)
output_png
```
