---
title: "PG_All Microsatellite Ordination Workflow"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(adegenet)
library(ggrepel)

report_dir <- dirname(knitr::current_input(dir = TRUE))
repo_root <- normalizePath(file.path(report_dir, "..", ".."), mustWork = FALSE)

input_file <- file.path(repo_root, "data", "PG_All.csv")
output_dir <- file.path(repo_root, "figure", "PG_All_ordination")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

state_colors <- c(
  "IN" = "#1b9e77",
  "KY" = "#d95f02",
  "EX-S" = "#7f7f7f",
  "TNE" = "#7570b3",
  "TNW" = "#e7298a",
  "Unknown" = "#666666"
)
```

## Goal

This report builds `adegenet`-based ordinations from the raw microsatellite matrix in `PG_All.csv`.

Requested outputs:

- PCA at the individual level
- PCA at the population level
- DAPC at the individual level
- DAPC at the population level

Important data note:

- `PG_All.csv` contains only `P. globosa` individuals
- a true all-species raw-genotype ordination cannot be produced from this file alone

## Read and Parse the Raw Microsatellite Matrix

```{r parse-raw}
raw <- readr::read_csv(input_file, col_names = FALSE, show_col_types = FALSE)

locus_names <- raw[3, -(1:2)] %>%
  unlist(use.names = FALSE) %>%
  as.character()
locus_names[is.na(locus_names)] <- ""

allele_names <- make.unique(locus_names[locus_names != ""])
allele_col_names <- paste0("allele_", seq_len(length(locus_names)))

sample_tbl <- raw[-c(1:3), ] %>%
  setNames(c("sample", "site", allele_col_names)) %>%
  mutate(
    sample = as.character(sample),
    site = as.character(site),
    site = str_replace(site, "^p", ""),
    population = str_extract(sample, "^Pg\\d+"),
    species = "P. globosa",
    state = case_when(
      population == "Pg346" ~ "IN",
      population %in% c("Pg347", "Pg349", "Pg350", "Pg351", "Pg352", "Pg354") ~ "KY",
      population %in% c("Pg348", "Pg353", "Pg355") ~ "EX-S",
      population %in% c("Pg356", "Pg357") ~ "TNE",
      population %in% c("Pg358", "Pg359", "Pg360", "Pg361", "Pg362") ~ "TNW",
      TRUE ~ "Unknown"
    )
  )

sample_tbl %>%
  select(sample, site, population, species, state) %>%
  slice_head(n = 10)
```

## Build Locus Genotypes for `adegenet`

```{r build-loci}
allele_cols <- allele_col_names

locus_tbl <- purrr::map_dfc(seq_along(allele_names), function(i) {
  allele_1 <- as.character(sample_tbl[[allele_cols[2 * i - 1]]])
  allele_2 <- as.character(sample_tbl[[allele_cols[2 * i]]])

  genotype <- if_else(
    is.na(allele_1) | is.na(allele_2) | allele_1 == "0" | allele_2 == "0",
    NA_character_,
    paste0(allele_1, "/", allele_2)
  )

  tibble(!!allele_names[i] := genotype)
})

locus_tbl %>% slice_head(n = 5)
```

## Create a `genind` Object

```{r genind}
genind_globosa <- adegenet::df2genind(
  locus_tbl,
  sep = "/",
  ploidy = 2,
  ind.names = sample_tbl$sample,
  pop = sample_tbl$population,
  NA.char = "NA",
  type = "codom"
)

genind_globosa
```

## PCA

```{r pca}
geno_tab <- adegenet::tab(genind_globosa, NA.method = "mean")
pca_obj <- prcomp(geno_tab, center = TRUE, scale. = FALSE)
explained <- (pca_obj$sdev^2) / sum(pca_obj$sdev^2)

pca_ind_scores <- as_tibble(pca_obj$x[, 1:2, drop = FALSE]) %>%
  setNames(c("Axis1", "Axis2")) %>%
  bind_cols(sample_tbl %>% select(sample, site, population, species, state))

pca_pop_scores <- pca_ind_scores %>%
  group_by(population, species, state) %>%
  summarise(Axis1 = mean(Axis1), Axis2 = mean(Axis2), n = n(), .groups = "drop")

pca_axis_labels <- c(
  sprintf("PC1 (%.1f%%)", 100 * explained[1]),
  sprintf("PC2 (%.1f%%)", 100 * explained[2])
)

pca_axis_labels
```

### PCA at the Individual Level

```{r pca-individual, fig.width=9, fig.height=7}
ggplot(pca_ind_scores, aes(x = Axis1, y = Axis2, color = state)) +
  geom_point(size = 2.3, alpha = 0.9) +
  stat_ellipse(aes(group = population), linewidth = 0.35, alpha = 0.35, show.legend = FALSE) +
  scale_color_manual(values = state_colors, drop = FALSE) +
  labs(
    title = "PCA at the individual level",
    subtitle = "PG_All.csv contains only P. globosa individuals",
    x = pca_axis_labels[1],
    y = pca_axis_labels[2],
    color = "State"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")
```

### PCA at the Population Level

```{r pca-population, fig.width=9, fig.height=7}
ggplot(pca_pop_scores, aes(x = Axis1, y = Axis2, label = population, color = state)) +
  geom_hline(yintercept = 0, color = "gray85", linewidth = 0.35) +
  geom_vline(xintercept = 0, color = "gray85", linewidth = 0.35) +
  geom_point(aes(size = n), alpha = 0.95) +
  geom_text_repel(size = 3.1, max.overlaps = 50) +
  scale_color_manual(values = state_colors, drop = FALSE) +
  scale_size_continuous(range = c(3, 8), name = "Individuals") +
  labs(
    title = "PCA at the population level",
    subtitle = "Population centroids from individual PCA scores",
    x = pca_axis_labels[1],
    y = pca_axis_labels[2],
    color = "State"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## DAPC

```{r dapc}
n_ind <- nInd(genind_globosa)
n_pop <- nPop(genind_globosa)
n_var <- ncol(adegenet::tab(genind_globosa, NA.method = "mean"))

n_pca <- max(2, min(40, n_ind - 1, n_var - 1))
n_da <- max(1, min(10, n_pop - 1))

dapc_obj <- adegenet::dapc(genind_globosa, pop(genind_globosa), n.pca = n_pca, n.da = n_da)

dapc_ind_scores <- as_tibble(dapc_obj$ind.coord[, 1:2, drop = FALSE]) %>%
  setNames(c("Axis1", "Axis2")) %>%
  bind_cols(sample_tbl %>% select(sample, site, population, species, state))

dapc_pop_scores <- dapc_ind_scores %>%
  group_by(population, species, state) %>%
  summarise(Axis1 = mean(Axis1), Axis2 = mean(Axis2), n = n(), .groups = "drop")

c(n_pca = n_pca, n_da = n_da)
```

### DAPC at the Individual Level

```{r dapc-individual, fig.width=9, fig.height=7}
ggplot(dapc_ind_scores, aes(x = Axis1, y = Axis2, color = state)) +
  geom_point(size = 2.3, alpha = 0.9) +
  stat_ellipse(aes(group = population), linewidth = 0.35, alpha = 0.35, show.legend = FALSE) +
  scale_color_manual(values = state_colors, drop = FALSE) +
  labs(
    title = "DAPC at the individual level",
    subtitle = sprintf("Retained %s PCs and %s discriminant axes", n_pca, n_da),
    x = "DA1",
    y = "DA2",
    color = "State"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")
```

### DAPC at the Population Level

```{r dapc-population, fig.width=9, fig.height=7}
ggplot(dapc_pop_scores, aes(x = Axis1, y = Axis2, label = population, color = state)) +
  geom_hline(yintercept = 0, color = "gray85", linewidth = 0.35) +
  geom_vline(xintercept = 0, color = "gray85", linewidth = 0.35) +
  geom_point(aes(size = n), alpha = 0.95) +
  geom_text_repel(size = 3.1, max.overlaps = 50) +
  scale_color_manual(values = state_colors, drop = FALSE) +
  scale_size_continuous(range = c(3, 8), name = "Individuals") +
  labs(
    title = "DAPC at the population level",
    subtitle = "Population centroids from individual DAPC scores",
    x = "DA1",
    y = "DA2",
    color = "State"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```
